Density-dependent selection and the limits of relative fitness

### 2018-02-27 manuscript.tex
l

The approximations \eqref{eq:meanfieldr} and \eqref{eq:meanfielda} must be done in a way that ensures consistency between rare and common types. To illustrate, suppose that two identical types are present (same $b$, $c$ and $d$), one with low density $l_1\ll 1$, and the other with high density $l_2\approx L\gg 1$. Since $L$ is large, uncontested territories contribute negligibly to each type's growth. The rare type grows almost entirely due to $\Delta_r n_1$, while the common type grows almost entirely due to $\Delta_a n_2$. To ensure consistency, the approximate per-capita growth rates $\Delta_r n_1/m_1 = \Delta_a n_2/m_2$ implied by \eqref{eq:meanfieldr} and \eqref{eq:meanfielda} must be equal. Even small violations of this consistency condition will cause one type to grow exponentially relative to the other. This is a logical contradiction: any single-type population can be divided at will into rare and common subsets of identical types, and clearly not all such subsets can grow/decline.

For example, suppose that we use $\tilde{p}$ and $\hat{p}$ in place of $\tilde{q}$ and $\hat{q}$ in Eqs.~\eqref{eq:meanfieldr} and \eqref{eq:meanfielda}, respectively. Then, since $\langle x_2\rangle_{\tilde{p}}\approx L$, the right hand side of Eq. \eqref{eq:meanfieldr} is approximately $1/(L+1)$ for the rare type. And, since $l_1\ll 1$ and $L\gg 1$ in Eq.~\eqref{eq:deltr}, we have $\Delta_r n_1 \approx 1/(L+1)$. Similarly, for the common type, $\sum_j \langle x_j\rangle_{\hat{p}}\approx L$ in Eq. \eqref{eq:meanfielda}, and so $\Delta_a n_2 \approx 1/L$. Therefore, the consistency requirement is violated if we naively use the conditional distributions $\tilde{p}$ and $\hat{p}$ derived from the dispersal distribution $p$ to calculate the effective means.

The effective distributions $\tilde{q}$ and $\hat{q}$ are chosen to ensure that this consistency requirement is satisfied.

>>>>

The approximations \eqref{eq:meanfieldr} and \eqref{eq:meanfielda} must be done in a way that ensures consistency between rare and common types. To illustrate, suppose that two identical (same $b$, $c$ and $d$) types are present, one with low density $l_1\ll 1$, and the other with high density $l_2\approx L\gg 1$. Since $L$ is large, uncontested territories contribute negligibly to each type's growth. The rare type grows almost entirely due to $\Delta_r n_1$, while the common type grows almost entirely due to $\Delta_a n_2$. To ensure consistency, the approximate per-capita growth rates $\Delta_r n_1/m_1 = \Delta_a n_2/m_2$ implied by the approximations \eqref{eq:meanfieldr} and \eqref{eq:meanfielda} must be equal. Even small violations of this consistency condition would mean exponential growth of one type relative to the other. This is a serious pathology: any single-type population can be arbitrarily partitioned into rare and common identical subtypes, and the predicted growth or decline of any small group of individuals would depend on whether it is regarded as part of a rare or a common type.

This pathology arises if we naively use $\tilde{p}$ and $\hat{p}$ in place of $\tilde{q}$ and $\hat{q}$ to calculate the effective means. In this case we have $\langle x_2\rangle_{\tilde{p}}\approx L$, the right hand side of Eq. \eqref{eq:meanfieldr} is approximately $1/(L+1)$ for the rare type, and, since $l_1\ll 1$ and $L\gg 1$ in Eq.~\eqref{eq:deltr}, we have $\Delta_r n_1 \approx 1/(L+1)$. Similarly, for the common type, $\sum_j \langle x_j\rangle_{\hat{p}}\approx L$ in Eq. \eqref{eq:meanfielda}, and so $\Delta_a n_2 \approx 1/L$. Thus, the identical rare type is erroneously predicted to decline in frequency.

The effective distributions $\tilde{q}$ and $\hat{q}$ ensure that the above consistency requirement is satisfied.
